Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-03-19 appendix.tex

\subsection{Radiation transport: ray tracing}
\label{sec.num.raytracing}

Local radiation from stars and black holes strongly affects their
surroundings.  Radiation transport is a well-studied problem; however,
its treatment in multidimensional calculations is difficult because of
the dependence on seven variables -- three spatial, two angular,
frequency, and time.  The non-local nature of the thermal and
hydrodynamical response to radiation sources further adds to the
difficulty.  Here we describe the ray tracing implementation \moray,
which is presented in full detail in \citet{Wise11_Moray} with seven
code tests and six applications.

The radiative transfer equation in comoving coordinates
\citep{Gnedin97} reads
%
\begin{equation}
  \label{eq:rteqn}
  \frac{1}{c} \; \frac{\partial I_\nu}{\partial t} + 
  \frac{\hat{n} \cdot \nabla I_\nu}{\bar{a}} -
  \frac{H}{c} \; \left( \nu \frac{\partial I_\nu}{\partial \nu} -
  3 I_\nu \right) = -\kappa_\nu I_\nu + j_\nu .
\end{equation}
%
Here $I_\nu \equiv I(\nu, \mathbf{x}, \Omega, t)$ is the radiation
specific intensity in units of energy per time $t$ per solid angle per
unit area per frequency $\nu$.  $H = \dot{a}/a$ is the Hubble
constant, where $a$ is the scale factor.  $\bar{a} = a/a_{em}$ is the
ratio of scale factors at the current time and time of emission.  The
second term represents the propagation of radiation, where the factor
$1/\bar{a}$ accounts for cosmic expansion.  The third term describes
both the cosmological redshift and dilution of radiation.  On the
right hand side, the first term considers the absorption coefficient
$\kappa_\nu \equiv \kappa_\nu(\mathbf{x},\nu,t)$, and the second term
$j_\nu \equiv j_\nu(\mathbf{x},\nu,t)$ is the emission coefficient
that includes any point sources of radiation or diffuse radiation.

We can make some appropriate approximations to reduce the complexity
of Equation (\ref{eq:rteqn}) in order to include radiation transport
in numerical calculations.  Typically timesteps in dynamic
calculations are small enough so that $\Delta a/a \ll 1$, therefore
$\bar{a} = 1$ in any given timestep, reducing the second term to
$\hat{n} \partial I_\nu/\partial \mathbf{x}$.  To determine the
importance of the third term, we evaluate the ratio of the third term
to the second term.  This is $HL/c$, where $L$ is the simulation box
length.  If this ratio is $\ll 1$, we can ignore the third term.  For
example at $z=5$, this ratio is 0.1 when $L = c/H(z=5)$ = 53 proper
Mpc.  In large boxes where the light crossing time is comparable to
the Hubble time, then it becomes important to consider cosmological
redshifting and dilution of the radiation.  Thus equation
(\ref{eq:rteqn}) reduces to the non-cosmological form in this local
approximation,
%
\begin{equation}
  \frac{1}{c} \frac{\partial I_\nu}{\partial t} + 
  \hat{n} \frac{\partial I_\nu}{\partial \mathbf{x}} =
  -\kappa_\nu I_\nu + j_\nu .
\end{equation}
%
We choose to represent the source term $j_\nu$ as point sources of
radiation (e.g. stars, quasars) that emit radial rays that are
propagated along the direction $\hat{n}$.

Ray tracing is an accurate method to propagate radiation from point
sources on a computational grid, given that there are a sufficient
number of rays passing through each cell.  Along a ray, the radiation
transfer equation reduces to
%
\begin{equation}
\label{eqn:rtray}
\frac{1}{c} \frac{\partial P}{\partial t} + \frac{\partial P}{\partial
  r} = -\kappa P,
\end{equation}
where $P$ is the photon number flux along the ray.  To sample the
radiation field at large radii, ray tracing requires at least $N_{ray}
= 4\pi R^2 / (\Delta x)^2$ rays to sample each cell with one ray,
where $R$ is the radius from the source to the cell and $\Delta x$ is
the cell width.  If one were to trace $N_{ray}$ rays out to $R$, the
cells at a smaller radius $r$ would be sampled by, on average,
$(r/R)^2$ rays, which is computationally wasteful because only a few
rays per cell, as we will show later, provides an accurate calculation
of the radiation field.

We avoid this inefficiency by utilizing adaptive ray tracing
\citep{Abel02_RT}, which progressively splits rays when the sampling
becomes too coarse and is based on Hierarchical Equal Area isoLatitude
Pixelation \citep[HEALPix;][]{HEALPix}.  In this scheme, the rays are
traced along normal directions of the centers of HEALPix pixels, which
evenly divides a sphere into equal areas.  The rays are initialized at
each point source with the photon luminosity (ph s$^{-1}$) equally
spread across $N_{\rm pix} = 12 \times 4^l$ rays, where $l$ is the
initial HEALPix level.  We usually find $l$ = 0 or 1 is sufficient
because these coarse rays will usually be split before traversing
the first cell.

The rays are traced through the grid in a typical fashion
\citep[e.g.][]{Abel99_RT}, in which we calculate the next cell
boundary crossing.  The ray segment length crossing the cell is
%
\begin{equation}
  \label{eqn:trace}
  dr = R_0 - \min_{i=1 \rightarrow 3} \left[(x_{\rm cell,i} - x_{\rm src,i}) /
    \hat{n}_{\rm ray,i} \right],
\end{equation}
%
where $R_0$, $\hat{n}_{\rm ray}$, $x_{\rm cell,i}$, and $x_{\rm
  src,i}$ are the initial distance traveled by the ray, normal
direction of the ray, the next cell boundary crossing in the $i$-th
dimension, and the position of the point source that emitted the ray,
respectively.  However before the ray travels across the cell, we
evaluate the ratio of the face area $A_{\rm cell}$ of the current cell
and the solid angle $\Omega_{\rm ray}$ of the ray,
%
\begin{equation}
  \label{eqn:split}
  \Phi_c = \frac{A_{\rm cell}} {\Omega_{\rm ray}} = 
  \frac{N_{\rm pix} (\Delta x)^2} {4\pi R_0^2}.
\end{equation}
%
If $\Phi_c$ is less than a pre-determined value (usually $>3$), the
ray is split into 4 child rays.  The pixel numbers of the child rays
$p^\prime$ are given by the ``nested'' scheme of HEALPix at the next
level, i.e. $p^\prime = 4 \times p + [0,1,2,3]$, where $p$ is the
original pixel number.  The child rays (1) acquire the new normal
vectors of the pixels, (2) retain the same radius of the parent ray,
and (3) gets a quarter of the photon flux of the parent ray.
Afterward the parent ray is discontinued.

A ray propagates and splits until (1) the photon has traveled $c
\times dt_P$, where $dt_P$ is the radiative transfer timestep; (2) its
photon flux is almost fully absorbed ($>99.9\%$) in a single cell,
which significantly reduces the computational time if the radiation
volume filling fraction is small; (3) the photon leaves the
computational domain with isolated boundary conditions; or (4) the
photon travels $\sqrt{3}$ of the simulation box length with periodic
boundary conditions.

In the first case, the photon is halted at that position and saved,
where it will be considered in the solution of $I_\nu$ at the next
timestep.  In the next timestep, the photon will encounter a different
hydrodynamical and ionization state, hence $\kappa$, in its path.
Furthermore any time variations of the luminosities will be retained
in the radiation field.  This is how this method retains the time
derivative of the radiative transfer equation.  The last restriction
prevents our method from considering sources external to the
computational domain, but a uniform radiation background can be used
in conjunction with ray tracing that adds the local radiation field to
the background intensity.

The radiation field is calculated by integrating equation
(\ref{eqn:rtray}) along each ray, which is done by considering the
discretization of the ray into segments.  In the following
description, we assume the rays are monochromatic for simplicity.  For
convenience, we express the integration in terms of optical depth
$\tau = \int \kappa(r,t) \; dr$, and for a ray segment,
%
\begin{equation}
  \label{eqn:dtau}
  d\tau = \sigma_{\rm abs}(\nu) n_{\rm abs} dr.
\end{equation}
Here $\sigma_{\rm abs}$ and $n_{\rm abs}$ are the cross section and
number density of the absorbing medium, respectively.  In the static
case, equation (\ref{eqn:rtray}) has a simple exponential analytic
solution, and the photon flux of a ray is reduced by
%
\begin{equation}
  \label{eqn:flux}
  dP = P \times (1 - e^{-\tau})
\end{equation}
as it crosses a cell.  We equate the photo-ionization rate to the
absorption rate, resulting in photon conservation \citep{Abel99_RT,
  Mellema06}.  Thus the photo-ionization $k_{\rm ph}$ and
photo-heating $\Gamma_{\rm ph}$ rates associated with a single ray are
%
\begin{equation}
  \label{eqn:kph}
  k_{\rm ph} = \frac{P (1 - e^{-\tau})}{n_{\rm abs} \; V_{\rm cell} \; dt_P},
\end{equation}
\begin{equation}
  \label{eqn:gamma}
  \Gamma_{\rm ph} = k_{\rm ph} \; (E_{\rm ph} - E_i),
\end{equation}
where $V_{\rm cell}$ is the cell volume, $E_{\rm ph}$ is the photon
energy, and $E_i$ is the ionization energy of the absorbing material.
In each cell, the photo-ionization and photo-heating rates from each
ray in the calculation are summed, and after the ray tracing is
complete, these rates are used to update the ionization state and
energy of the cells with the solver described in \S\ref{sec.ov.chem}.

>>>>

\appendix
\section{Interpolation methods}
\label{app:interpolation}

### 2013-03-19 code-tests-fld.tex

This test problem examines Enzo's FLD radiation transport and
associated ionization chemistry solvers on a cosmological I-front
test.  This corresponds to the test problem 4.6
from \cite{ReynoldsHayesPaschosNorman2009}, run using the cosmological
deceleration parameter $q_0 = 0.05$ and initial redshift $z_i=10$.
Here, the physics of interest is the expansion of a HII region in a
uniform gas around a single monochromatic ionizing source (with
frequency $h\nu = 13.6$ eV).  In this problem the I-front propagates
rapidly at first, approaching 90\% of the Str{\" o}mgren radius by a
scaled redshift of $-\log\left[(1+z)/(1+z_i)\right] \approx 0.15$,
before cosmological expansion overtakes ionization, pushing the Str{\"
o}mgren radius outward faster than the I-front can propagate.  Here,
the Str{\" o}mgren radius is analytically given by
\[
   r_S(t) = \left[\frac{3\dot{N}_{\gamma}}{4\pi \alpha_B
   n_H(t)^2}\right]^{1/3}, 
\]
where $\dot{N}_{\gamma} = 5\cdot10^{48}$ photon sec$^{-1}$ is the
stength of the ionizing source, $\alpha_B = 2.52\cdot10^{-13}$
s$^{-1}$ is the HII recombination rate, and $n_H(t)$ is the proper
number density of hydrogen.  We then define $\lambda = \alpha_B\,
n_{H,i}\, /\, H_0\, / (1+z_i)$, where the subscript $i$ indicates the
quantity at the initial redshift.  The analytical solution for the
I-front position as a function of time (or actually of the
time/redshift-dependent expansion coefficient $a(t)$) is then
\begin{eqnarray*}
   r_I(t) &=& r_{S,i} \left[\lambda e^{-\tau(a)} \int_1^{a(t)}
     e^{\tau(\tilde a)} \left(1 - 2q_0
     + \frac{2q_0(1+z_i)}{\tilde{a}}\right)^{-1/2}\mathrm
     d\tilde{a}\right]^{1/3}, \quad\text{where} \\ 
   \tau(a) &=& \lambda\frac{6q_0^2(1+z_i)^2}{F(a)-F(1)}, 
     \quad\text{and} \\
   F(a) &=& \left[2-4q_0 - \frac{2q_o(1+z_i)}{a}\right] 
      \left[1-2q_0 + \frac{2q_0(1+z_i)}{a}\right]^{1/2}.
\end{eqnarray*}
The simulation parameters are a box size $L_i\approx 27$ kpc, Hubble
constant $H_0 = 1$, and cosmological constants $\Omega_m = 0.1$,
$\Omega_A=0$ and $\Omega_b = 0.1$.  The initial values for the
simulation are a radiation energy density $E_{r,i} = 10^{-35}$ erg
cm$^{-3}$, temperature $T_i = 10^4$ K, density $\rho_{b,i} =
2.35\cdot10^{-28}$ g cm$^{-3}$, and an ionized hydrogen fraction
${\mathrm n}_{HII}/{\mathrm n}_H = 0$.

In the Figure \ref{fig.fld}, we plot profiles of the HI and HII
fractions at the redshifts $z=\{6.24,\, 2.29,\, 1.02\}$, as well as the
ratio of the I-front and Str{\" o}mgren radii throughout the
simulation.  As seen in the profile figure, the I-front width starts
out quite sharp but slowly spreads as the simulation proceeds due to
the diffusion approximation to the radiative transfer.  However, as
seen in the I-front propagation history plot, the computed I-front
location very accurately captures the exact value.

>>>>

This test problem examines Enzo's FLD radiation transport and
associated ionization chemistry solvers on a cosmological I-front
test.  This corresponds to the test problem 4.6
from \cite{ReynoldsHayesPaschosNorman2009}, run using the cosmological
deceleration parameter $q_0 = 0.05$ and initial redshift $z_i=10$.
Here, the physics of interest is the expansion of a \ion{H}{2} region in a
uniform gas around a single monochromatic ionizing source (with
frequency $h\nu = 13.6$ eV).  In this problem the I-front propagates
rapidly at first, approaching 90\% of the Str{\" o}mgren radius by a
scaled redshift of $-\log\left[(1+z)/(1+z_i)\right] \approx 0.15$,
before cosmological expansion overtakes ionization, pushing the Str{\"
o}mgren radius outward faster than the I-front can propagate.  Here,
the Str{\" o}mgren radius is analytically given by
\[
   r_S(t) = \left[\frac{3\dot{N}_{\gamma}}{4\pi \alpha_{\rm B}
   n_{\rm H}(t)^2}\right]^{1/3}, 
\]
where $\dot{N}_{\gamma} = 5\cdot10^{48}$ photon sec$^{-1}$ is the
stength of the ionizing source, $\alpha_{\rm B} = 2.52\cdot10^{-13}$
s$^{-1}$ is the \ion{H}{2} recombination rate, and $n_H(t)$ is the
proper number density of hydrogen.  We then define $\lambda =
\alpha_{\rm B}\, n_{H,i}\, /\, H_0\, / (1+z_i)$, where the subscript $i$
indicates the quantity at the initial redshift.  The analytical
solution for the I-front position as a function of time (or actually
of the time/redshift-dependent expansion coefficient $a(t)$) is then
\begin{eqnarray*}
   r_I(t) &=& r_{S,i} \left[\lambda e^{-\tau(a)} \int_1^{a(t)}
     e^{\tau(\tilde a)} \left(1 - 2q_0
     + \frac{2q_0(1+z_i)}{\tilde{a}}\right)^{-1/2}\mathrm
     d\tilde{a}\right]^{1/3}, \quad\text{where} \\ 
   \tau(a) &=& \lambda\frac{6q_0^2(1+z_i)^2}{F(a)-F(1)}, 
     \quad\text{and} \\
   F(a) &=& \left[2-4q_0 - \frac{2q_o(1+z_i)}{a}\right] 
      \left[1-2q_0 + \frac{2q_0(1+z_i)}{a}\right]^{1/2}.
\end{eqnarray*}
The simulation parameters are a box size $L_i\approx 27$ kpc, Hubble
constant $H_0 = 1$, and cosmological constants $\Omega_m = 0.1$,
$\Omega_A=0$ and $\Omega_b = 0.1$.  The initial values for the
simulation are a radiation energy density $E_{r,i} = 10^{-35}$ erg
cm$^{-3}$, temperature $T_i = 10^4$ K, density $\rho_{b,i} =
2.35\cdot10^{-28}$ g cm$^{-3}$, and an ionized hydrogen fraction
$n_{\rm HII}/n_{\rm H} = 0$.

In the Figure \ref{fig.fld}, we plot profiles of the \ion{H}{1} and
\ion{H}{2} fractions at the redshifts $z=\{6.24,\, 2.29,\, 1.02\}$, as
well as the ratio of the I-front and Str{\" o}mgren radii throughout
the simulation.  As seen in the profile figure, the I-front width
starts out quite sharp but slowly spreads as the simulation proceeds
due to the diffusion approximation to the radiative transfer.
However, as seen in the I-front propagation history plot, the computed
I-front location very accurately captures the exact value.

### 2013-03-19 numerical-amr.tex

In one dimension, such functions are well known; examples include
donor cell, Van-Leer and piecewise parabolic interpolation, which are
first, second and third-order accurate respectively \citep[e.g.,][]{Stone92a}.
However, we require fully multi-dimensional interpolation functions, and so
focus on second-order schemes since they are tractable and
relatively accurate. 
\footnote{We could, in principle, build a multi-dimensional interpolator out
of a one-dimensional interpolation function; however, at higher order, this would not necessarily
satisfy our constraints on not introducing new extrema.}
In Appendix~\ref{app:interpolation} we explicitely write down
functions in one, two and three dimensions, but here we discuss the
general design considerations.

>>>>

In one dimension, such functions are well known; examples include
donor cell, Van-Leer and piecewise parabolic interpolation, which are
first, second and third-order accurate respectively
\citep[e.g.,][]{Stone92a}.  However, we require fully
multi-dimensional interpolation functions, and so focus on
second-order schemes since they are tractable and relatively accurate.
\footnote{We could, in principle, build a multi-dimensional
  interpolator out of a one-dimensional interpolation function;
  however, at higher order, this would not necessarily satisfy our
  constraints on not introducing new extrema.}  In the Appendix, we
explicitely write down functions in one, two and three dimensions, but
here we discuss the general design considerations.

### 2013-03-19 numerical-chemistry.tex

The primordial non-equilibrium chemistry routines used in \enzo\ were
first described by Abel et al. and Anninos et
al. \citep{abel97,anninos97}, but have been extended since then with
updated reaction rates and the inclusion of deuterium species
\citep{2009PhDT.........5T}.  These routines follow the
non-equilibrium chemistry of a gas of primordial composition with 12
total species: $H$, $H^+$, $He$, $He^+$, $He^{++}$, $H^-$, $H_2^+$,
$H_2$, $e^-$, $D$, $D^+$, and $HD$.  (A separate, but related, module
follows the radiative heating and cooling of the gas from atomic and
molecular line excitation, recombination, collisional excitation,
free-free transitions, Compton scattering of the cosmic microwave
background, as well as several models for a metagalactic UV background
that heat the gas via photoionization and discussion -- see
Section~\ref{sec.num.cooling} for more details).  Depending on user
parameters, one can follow simply atomic species ($H$, $H^+$, $He$,
$He^+$, $He^{++}$, and $e^-$), include those relevant for molecular
hydrogen formation ($H_2$, $H_2^+$, and $H^-$), and further include
deuterium and its species ($D$, $D^+$, and $HD$.)  \red{A total of 9
kinetic equations are solved, including 29 kinetic and radiative
processes, for the 9 species mentioned above.  See
Table~\ref{table.collisional} for the collisional processes and
Table~\ref{table.radiative} for the radiative processes solved.}

@@

\enzo\ solves the rate equations using a method based on a backwards
differencing formula (BDF) in order to provide a stable and accurate
solution. This technique is optimized by taking the chemical
intermediaries $H^-$ and $H_2^+$, which have large rate coefficients
and low concentrations, and grouping them into a separate category of
equations.  Due to their fast reactions these species are very
sensitive to small changes in the more abundant species.  However, due
to their low overall concentrations, they do not significantly
influence the abundance of species with higher concentrations.
Therefore, reactions involving these two species can be decoupled from
the rest of the network and treated independently.  In fact, $H^-$ and
$H_2^+$ are treated as being in equilibrium at all times, independent
of the other species and the hydrodynamic variables.  This allows a
large speedup in solution as the BDF scheme is then applied only to
the slower 7- or 10-species network (depending on whether deuterium is
included or not) on timescales closer to those required by the
hydrodynamics of the simulation.  Even so, the accuracy and stability
of the scheme is maintained by subcycling the rate solver over a
single hydrodynamic timestep.  These subcycle timesteps are determined
so that the maximum fractional change in the electron concentration is
limited to no more than $10\%$ per timestep.  Given this subcycling,
the method is not truly implicit over an entire hydrodynamical
timestep, and thus not truly a BDF method -- however, it uses the
formalism and is implicit over subcycle timesteps.

It is important to note the regime in which this model is valid.
According to Abel et al. and Anninos et al. \citep{abel97,anninos97},
the reaction network is valid for temperatures between $10^0 - 10^8$
K.  The original model discussed in these two references is only valid
up to $n_H \sim 10^4$~cm$^{-3}$.  However, addition of the 3-body
$H_2$ formation process (equation 20 in Table~\ref{table.collisional})
allows correct modeling of the chemistry of the gas up until the point
where collisionally-induced emission from molecular hydrogen becomes
an important cooling process, which occurs at $n_H \sim
10^{14}$~cm$^{-3}$.  A further concern is that the optically thin
approximation for radiative cooling breaks down, which occurs before
$n_H \sim 10^{16} - 10^{17}$~cm$^{-3}$.  Beyond this point,
modifications the cooling function that take into account the
non-negligible opacity in the gas must be made, as discussed by
\citet{2004MNRAS.348.1019R}, and was put into \enzo\ for the work
published in \citep{2009Sci...325..601T,2009PhDT.........5T}.  Even with these
modifications, a completely correct description of the cooling of this
gas will require some form of radiation transport, which will greatly
increase the cost of the simulations.

>>>>

The primordial non-equilibrium chemistry routines used in \enzo\ were
first described by Abel et al. and Anninos et
al. \citep{abel97,anninos97}, but have been extended since then with
updated reaction rates and the inclusion of deuterium species
\citep{2009PhDT.........5T}.  These routines follow the
non-equilibrium chemistry of a gas of primordial composition with 12
total species: H, \Hp, He, \Hep, \Hepp, \Hm, \HHp, \HH, e$^-$, D, \Dp,
and HD.  (A separate, but related, module follows the radiative
heating and cooling of the gas from atomic and molecular line
excitation, recombination, collisional excitation, free-free
transitions, Compton scattering of the cosmic microwave background, as
well as several models for a metagalactic UV background that heat the
gas via photoionization and discussion -- see
Section~\ref{sec.num.cooling} for more details).  Depending on user
parameters, one can follow simply atomic species (H, \Hp, He, \Hep,
\Hepp, and e$^-$), include those relevant for molecular hydrogen
formation (\HH, \HHp, and \Hm), and further include deuterium and its
species (D, \Dp, and HD).  \red{A total of 9 kinetic equations are
  solved, including 29 kinetic and radiative processes, for the 9
  species mentioned above.  See Table~\ref{table.collisional} for the
  collisional processes and Table~\ref{table.radiative} for the
  radiative processes solved.}

@@

\enzo\ solves the rate equations using a method based on a backwards
differencing formula (BDF) in order to provide a stable and accurate
solution. This technique is optimized by taking the chemical
intermediaries \Hm and \HHp, which have large rate coefficients
and low concentrations, and grouping them into a separate category of
equations.  Due to their fast reactions these species are very
sensitive to small changes in the more abundant species.  However, due
to their low overall concentrations, they do not significantly
influence the abundance of species with higher concentrations.
Therefore, reactions involving these two species can be decoupled from
the rest of the network and treated independently.  In fact, \Hm~and
\HHp~are treated as being in equilibrium at all times, independent
of the other species and the hydrodynamic variables.  This allows a
large speedup in solution as the BDF scheme is then applied only to
the slower 7- or 10-species network (depending on whether deuterium is
included or not) on timescales closer to those required by the
hydrodynamics of the simulation.  Even so, the accuracy and stability
of the scheme is maintained by subcycling the rate solver over a
single hydrodynamic timestep.  These subcycle timesteps are determined
so that the maximum fractional change in the electron concentration is
limited to no more than $10\%$ per timestep.  Given this subcycling,
the method is not truly implicit over an entire hydrodynamical
timestep, and thus not truly a BDF method -- however, it uses the
formalism and is implicit over subcycle timesteps.

It is important to note the regime in which this model is valid.
According to \citet{abel97} and \citet{anninos97},
the reaction network is valid for temperatures between $10^0 - 10^8$
K.  The original model discussed in these two references is only valid
up to $n_H \sim 10^4$~cm$^{-3}$.  However, addition of the 3-body
\HH~formation process (equation 20 in Table~\ref{table.collisional})
allows correct modeling of the chemistry of the gas up until the point
where collisionally-induced emission from molecular hydrogen becomes
an important cooling process, which occurs at $n_{\rm H} \sim
10^{14}$~cm$^{-3}$.  A further concern is that the optically thin
approximation for radiative cooling breaks down, which occurs before
$n_{\rm H} \sim 10^{16} - 10^{17}$~cm$^{-3}$.  Beyond this point,
modifications the cooling function that take into account the
non-negligible opacity in the gas must be made, as discussed by
\citet{2004MNRAS.348.1019R}, and was put into \enzo\ for the work
published in \citep{2009Sci...325..601T,2009PhDT.........5T}.  Even
with these modifications, a completely correct description of the
cooling of this gas will require some form of radiation transport,
which will greatly increase the cost of the simulations.

### 2013-03-19 numerical-cooling.tex

\enzo has multiple methods for computing the energy change from
radiative cooling and heating.  All of them assume the gas is totally
optically thin.  Below we describe the methods for computing the
cooling rates from metal-free and metal-enriched gas.  Sample cooling
curves for each of \enzo's primary cooling methods are shown in Figure
\ref{fig.cooling_rate}.

@@

As discussed in \S\ref{sec.ov.chem}, the set of reactions that
characterize a metal-free gas is simple enough to be computed in
non-equilibrium during the simulation.  Similarly, the radiative
cooling of metal-free gas is solved by directly computing the cooling
and heating rates from the following individual processes for atomic
H and He: collisional excitation and ionization, recombination,
free-free emission, Compton scattering off of the cosmic microwave
background (CMB), and photo-heating from a UV metagalactic background.
If the H$_{2}$ chemistry network is enabled, the following H$_{2}$
cooling processes are also considered: ro-vibrational transitions 
\citep{1998A&A...335..403G}, heating and cooling from molecular
formation and destruction \citep{2009Sci...325..601T}, and 
collision-induced emission \citep{2004MNRAS.348.1019R}.  If Deuterium
chemistry is enabled, then rotational transitions of HD
\citep{1998A&A...335..403G} are treated as well.  The radiative
cooling calculation is coupled to the update of the chemistry network
such that they both occur within the same subcycling loop.  This is
necessary in regimes where rapid cooling or change in the ionization
state occur.  In
addition to the subcycle timestepping contraints mentioned in
\S\ref{sec.ov.chem}, the subcycle timestep is also not permitted to
exceed 10\% of the cooling time, $e/\dot{e}$.  A metagalactic
background affects the gas through both photo-heating and
photo-ionization.  These are treated by including redshift-dependent 
photo-ionization and photo-heating rate terms in the chemistry and
cooling equations for HI, HeI, and HeII.  The black curves in Figure
\ref{fig.cooling_rate} show cooling rates for a metal-free gas with
number density $n_{H} = 10^{-4}$ cm$^{-3}$ both with and without a
radiation background.  More detail on the specific UV backgrounds
present in \enzo is presented below in
\S\ref{sec.num.rad-homogeneous}.

@@

A more sophisticated method makes use of multidimensional cooling and
heating rate tables computed with the photo-ionization code,
\texttt{Cloudy} \citep{1998PASP..110..761F}.  This method, detailed in
\citet{2008MNRAS.385.1443S} and \citet{2011ApJ...731....6S}, works by
using \texttt{Cloudy} to compute the cooling and heating rates from
the metal species only.  The primary assumption made is that of
ionization equilibrium.  The tables can have up to five dimensions,
where those dimensions are density, metallicity, electron fraction,
redshift (for an evolving metagalactic UV backgroun), and
temperature.  Tables can be created for any abundance pattern for
elements up to atomic number 30 (Zn) and for any incident radiation
field.  The cooling from H and He is calculated by the non-equilibrium
primordial chemistry and cooling solver.  The contribution from metals
to the cooling is computed within the same subcycling loop as the
coupled primordial chemistry and cooling solver.  The blue curves in
Figure \ref{fig.cooling_rate} show cooling rates calculated with this
method for a gas with number density $n_{H} = 10^{-4}$ cm$^{-3}$ and
metallicity of 0.5 Z$_{\odot}$.

>>>>

\enzo\ has multiple methods for computing the energy change from
radiative cooling and heating.  All of them assume the gas is totally
optically thin.  Below we describe the methods for computing the
cooling rates from metal-free and metal-enriched gas.  Sample cooling
curves for each of \enzo's primary cooling methods are shown in Figure
\ref{fig.cooling_rate}.

@@

As discussed in \S\ref{sec.ov.chem}, the set of reactions that
characterize a metal-free gas is simple enough to be computed in
non-equilibrium during the simulation.  Similarly, the radiative
cooling of metal-free gas is solved by directly computing the cooling
and heating rates from the following individual processes for atomic H
and He: collisional excitation and ionization, recombination,
free-free emission, Compton scattering off of the cosmic microwave
background (CMB), and photo-heating from a UV metagalactic background.
If the H$_{2}$ chemistry network is enabled, the following H$_{2}$
cooling processes are also considered: ro-vibrational transitions
\citep{1998A&A...335..403G}, heating and cooling from molecular
formation and destruction \citep{2009Sci...325..601T}, and
collision-induced emission \citep{2004MNRAS.348.1019R}.  If Deuterium
chemistry is enabled, then rotational transitions of HD
\citep{1998A&A...335..403G} are treated as well.  The radiative
cooling calculation is coupled to the update of the chemistry network
such that they both occur within the same subcycling loop.  This is
necessary in regimes where rapid cooling or change in the ionization
state occur.  In addition to the subcycle timestepping contraints
mentioned in \S\ref{sec.ov.chem}, the subcycle timestep is also not
permitted to exceed 10\% of the cooling time, $e/\dot{e}$.  A
metagalactic background affects the gas through both photo-heating and
photo-ionization.  These are treated by including redshift-dependent
photo-ionization and photo-heating rate terms in the chemistry and
cooling equations for \ion{H}{1}, \ion{He}{1}, and \ion{He}{2}.  The
black curves in Figure \ref{fig.cooling_rate} show cooling rates for a
metal-free gas with number density $n_{\rm H} = 10^{-4}$ cm$^{-3}$ both
with and without a radiation background.  More detail on the specific
UV backgrounds present in \enzo is presented below in
\S\ref{sec.num.rad-homogeneous}.

@@

A more sophisticated method makes use of multidimensional cooling and
heating rate tables computed with the photo-ionization code,
\texttt{Cloudy} \citep{1998PASP..110..761F}.  This method, detailed in
\citet{2008MNRAS.385.1443S} and \citet{2011ApJ...731....6S}, works by
using \texttt{Cloudy} to compute the cooling and heating rates from
the metal species only.  The primary assumption made is that of
ionization equilibrium.  The tables can have up to five dimensions,
where those dimensions are density, metallicity, electron fraction,
redshift (for an evolving metagalactic UV backgroun), and
temperature.  Tables can be created for any abundance pattern for
elements up to atomic number 30 (Zn) and for any incident radiation
field.  The cooling from H and He is calculated by the non-equilibrium
primordial chemistry and cooling solver.  The contribution from metals
to the cooling is computed within the same subcycling loop as the
coupled primordial chemistry and cooling solver.  The blue curves in
Figure \ref{fig.cooling_rate} show cooling rates calculated with this
method for a gas with number density $n_{\rm H} = 10^{-4}$ cm$^{-3}$ and
metallicity of 0.5 Z$_{\odot}$.

### 2013-03-19 numerical-particle.tex
Adding ray tracing method section (a little more than a page).  Fixed a lot of missing references and citations.

The codes needs to be able to model collisionless 
One way would be to model the dark matter (or other collisionless particle-like objects, such as stars) as a second fluid in addition to the baryonic fluid and solve the collisionless Boltzmann equation, which follows the evolution of the fluid density in six-dimensional phase space.  However, this is computationally prohibitive owing to the large dimensionality of the problem, making this approach  unappealing for the cosmological AMR code.
%GB: removed the above (good discussion but not really appropriate in this paper I think, or maybe at the beginning)

To solve the, \enzo\ uses a particle-mesh N-body method to calculate 
the dynamics of collisionless systems \citep{Hockney88}.  This method 
follows trajectories 
of a representative sample of individual particles and is much more 
efficient than a direct solution of the Boltzmann equation in most 
astrophysical situations. 
The gravitational potential is computed by solving the elliptic 
Poisson's equation (Eqtn.~\ref{enzopoisson}).

\begin{eqnarray}
x^{n+1/2} & = & x^n + (\Delta t/2) v^{n} \nonumber \\
v^{n+1} & = & v^n + \Delta t a^{n+1/2} \\
x^{n+1} & = & x^{n+1/2} + (\Delta t/2) v^{n+1} \nonumber
\end{eqnarray}

>>>>

% The code needs to be able to model collisionless matter.  One way
% would be to model the dark matter (or other collisionless
% particle-like objects, such as stars) as a second fluid in addition to
% the baryonic fluid and solve the collisionless Boltzmann equation,
% which follows the evolution of the fluid density in six-dimensional
% phase space.  However, this is computationally prohibitive owing to
% the large dimensionality of the problem, making this approach
% unappealing for the cosmological AMR code.
%GB: removed the above (good discussion but not really appropriate in this paper I think, or maybe at the beginning)

To solve the, \enzo\ uses a particle-mesh N-body method to calculate 
the dynamics of collisionless systems \citep{Hockney88}.  This method 
follows trajectories 
of a representative sample of individual particles and is much more 
efficient than a direct solution of the Boltzmann equation in most 
astrophysical situations. 
The gravitational potential is computed by solving the elliptic 
Poisson's equation (Eq.~\ref{eq:potential}).

\begin{eqnarray}
\label{eqn.driftkick}
x^{n+1/2} & = & x^n + (\Delta t/2) v^{n} \nonumber \\
v^{n+1} & = & v^n + \Delta t a^{n+1/2} \\
x^{n+1} & = & x^{n+1/2} + (\Delta t/2) v^{n+1} \nonumber
\end{eqnarray}
